---
title: "Load packages and data"
format:
  html:
    self-contained: true
    code-copy: true

execute:
  eval: false
  message: false
  warning: false
---
### Generic stuff

#### Load packages (install them first if aren't already)

The packages `ggplot2` and `patchwork` are essential for these plots. The other packages are used a bit here and there. The individual packages will be mentioned the first time they are used.

```{r}
library(ggplot2)
library(patchwork)
library(data.table)
library(dplyr)
library(pammtools)
library(scales)
```

#### Import data

We load data with the function `fread()` (**f**ast **read**) from the mighty `data.table` package. We do a bit of data management with functions from the very user-friendly (but less efficient) package `dplyr`.

Below, we set labels for the exposure variable, you may use different labels if you have other preferences.

The data set contains results for different analyses, presently we only need data for overall survival. We filter (restrict) to get the relevant data.

```{r}
survival <- fread("//uni.au.dk/dfs/HE_KEA-DCESAS02/S/Stat day May 2025/risk_curves/data/riskdata.csv") |>
  mutate(x = factor(
    x,
    levels = c(2,1),
    labels = c("Control", "Intervention")
  )) |>
  filter(analysis == "Survival")

```

#### Global values

Set preferences for base size (this will affect sizes of axis titles and labels, legends, etc. but not the size of elements within the plot) and specify your 2 favorite colors for plotting curves. These preferences will be used in all plots.

Base size:

```{r}
bsize <- 15
```

There are many different [named colors](https://debruine.github.io/30DCC-2022/app-a-colours.html), e.g:

```{r}
mypal <- c("navajowhite", "firebrick")
```

You can use hex codes if you want to be specific about which colors to use:

```{r}
mypal <- c("#aaaaaa", "#af0000")
```

Whenever we need to set the base size or specify colors in the following, we can use `bsize` and `mypal`. The upshot of specifying these values is that if we want to change the base size or the colors, we need to do so only once.