---
title: "A 'Risk table' function"
format:
  html:
    self-contained: true
    code-copy: true

execute:
  eval: false
  message: false
  warning: false
---

### Avoid repetitions - use functions (again)

We will now make a function that can produce risk tables for a each level of `analysis`, as we did for the risk curves.

Again, it is probably a good idea to make a risk table within one level of `analysis` and then wrap it up inside a function.

As for the risk curve function, you might want to

-   start by subsetting the data (`atrisk_risk`) to a level of `analysis`,

-   develop a risk table as you did for the first exercise

    -   Consider if you want to have colors in this risk table - we have a legend inside the plot. Do what you find aesthetically most pleasing,

-    wrap it up into a function (you can call it `atrisk_fun`).

::: {.callout-tip collapse="true" appearance="simple"}
## Solution

```{r}
atrisk_fun <- function(thisanalysis){
  temp <- atrisk_risk |>
    filter(analysis == thisanalysis)
  ggplot(temp, aes(x = time, y = x, label = natrisk)) +
    geom_text(size = 4) +
    theme_void(base_size = bsize) +
    theme(axis.text.y = element_text(hjust = 1),
          plot.title = element_text(hjust = -.1, size = bsize)) +
    ggtitle("At risk")
}
```
:::

::: {.callout-note collapse="true" appearance="default" icon="false"}
## The risk table for `"Cholera"`

```{r}
atrisk_fun("Cholera")
```


```{r, echo = FALSE, eval = TRUE}
source("xx_generics.r")

atrisk_fun <- function(thisanalysis){
  temp <- atrisk_risk |>
    filter(analysis == thisanalysis)
  ggplot(temp, aes(x = time, y = x, label = natrisk)) +
    geom_text(size = 4) +
    theme_void(base_size = bsize) +
    theme(axis.text.y = element_text(hjust = 1),
          plot.title = element_text(hjust = -.1, size = bsize)) +
    ggtitle("At risk")
}

atrisk_fun("Cholera")
```
:::
